A Framework for Discrete Change

### 2008-11-10 papers/individual/individual.tex
added new bib file and updated refs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@238 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

% phenomena of discrete change in development/learning
Piagetian developmental theory (REFERENCE??)  assumes step-wise
changes in the strategies that children apply in all kinds of tasks
such as the conservation learning and the balance scale task.  Van der
Maas et al.  (1992) developed a catastrophe model to describe phase
transitions in learning and developmental processes.  They applied the
catastrophe model to learning in the conservation of liquid task
(REFERENCE??)  in which children have to judge relative volumes of
liquid in glasses of different heights and widths.  Young children
tend to ignore the width dimension and hence always choose the glass
with the highest level of liquid (REFERENCE??).  Van der Maas et al.
(1992) showed that there is a sudden transition to a new strategy in
which the children also take the width of the glasses into account
when judging the volume of liquids.

% balance scale rules
Jansen and Van der Maas (2001) applied the catastrophe model to
development of strategies on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 (Siegler, 1981).  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 (Siegler, 1981).

% hysteresis on the balance scale
Jansen and Van der Maas (2001) found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, they found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  Jansen and Van der Maas also found
evidence for hysteresis: the phenomenon that switching between
strategies is assymetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior (Gallistel et al 2004).  In particular,
in their study, evidence was found for sudden onset of learning: at
the start of the learning experiment, the pigeons did not learn
anything and performance was stable; after a number of trials,
learning kicks in and there are large increases in performance.  The
interest here is in modeling the distribution of onset times: that is,
the trials at which learning suddenly takes off.  A similar interest
in process onset times is found in addiction research.  For example,
(REFERENCE) study the age at which children start using alcohol and
how this related to eventual outcomes in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  This kind of learning is referred to as all-or-none
learning or concept identification learning.  Raijmakers et al (2001)
found evidence for different strategies applied by children when faced
with such a learning task.  Schmittmann et al (2006) reanalyzed their
data using hidden Markov models to show that both strategies are
characterized by sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example, Visser,
Raijmakers, \& Van der Maas (2008) analyzed data from a single
participant in an experimental task that manipulates the trade-off
between speed and accuracy.  The data consisted of three time series
with each around 150 repeated measurements of both reaction time and
accuracy.  Below we provide examples of analyzing time series from
single participants from two experiments; one from the Iowa Gambling
Task and one from the weather prediction task.  The interest in these
tasks is to show that participants develop different strategies over
time in responding to the stimuli and that the transition from one
strategy to the next is a discrete event.  Before providing these
illustrations, below we give a formalization of dependent mixture
models and a brief overview of the DepmixS4 package that was developed
to specify and fit such models.

@@

% what is the IGT?
The Iowa gambling task (IGT) is an experimental paradigm designed to
mimic real-life decision-making situations (Bechara, Damasio, Damasio
\& Anderson, 1994), in the way that it factors uncertainty, reward and
punishment (Dunn, Dalgleish, \& Lawrence, 2006).  The task requires
the selection of cards from four decks.  Each deck is characterized by
a certain amount of gain (delivered on each draw), frequency of loss,
and amount of loss.  Two decks (A and B) yield consistently high
rewards, but also high, probabilistic penalties and are both (equally)
disadvantageous in the long run.  The other two decks (C and D) yield
consistently smaller rewards, but also low, probabilistic penalties
and are both (equally) advantageous in the long run.  It is assumed
that the ventromedial prefrontal cortex (VMPFC) is active in the IGT
as VMPC patients show impaired task performance.  Their preference for
the decks with immediate high rewards indicates ``myopia for the
future''.

% what is the HDT? developmental trends and relevance?
Crone \& van der Molen (2004) designed a developmentally appropriate
analogue of the IGT, the Hungry Donkey Task (HDT), with a similar win
and loss schedule although the abolute amounts were redcuced by a
factor of 25.  The HDT is a pro-social game inviting the player to
assist a hungry donkey to collect as many apples as possible, by
opening one of four doors.  Again, doors A and B are characterized by
a high constant gain (10 apples), whereas doors C and D deliver a low
constant gain (2 apples).  At doors A and C, a loss of 50 apples (A)
or 10 apples (C) is delivered in 50\% of the trials.  For doors B and
D, frequency of loss is only 10\%.  The median loss of doors B and D
is 10 and 2, respectively.  Crone and van der Molen administered the
HDT to children from four age groups (6-9, 10-12, 13-15, and 18-25
year-olds) and concluded that children also fail to consider future
consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset (Huizenga, Crone, \& Jansen, 2007)
indicated that participants might solve the task by sequentially
considering the three dimensions (constant gain, frequency of loss,
and amount of loss) in order to choose a door.  Most youngest children
in the dataset seem to focus on the dominant dimension in the task,
frequency of loss, resulting in equal preference for doors B and D.
Older participants seem to use a two-dimensional rule where
participants first focus on the frequency of loss and then consider
amount of loss, resulting in a preference for door D. A third very
small subgroup seems to use an integrative rule where participants
combine all three dimensions in the appropriate way.  Participants
using the integrative rule pick cards from doors C and D, which are
advantageous in the long run.

@@

We chose to analyze two participants that used different strategies at
the end of the task as analyzed in the manner proposed by Huizenga et
al\.  (2007).  The first participants' data were best described by a
4-state model.  This model's transition matrix is: 
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbind.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~ref{tab:igt4} below.

@@

\begin{figure}
\begin{center}
	\scalebox{0.9}{\includegraphics*{graphs/post4.pdf}}
	\label{fig:post4}
	\caption{Posterior state sequence for the 4 state model.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
	\scalebox{1}{\includegraphics*{graphs/igt4.pdf}}
	\label{fig:igt4}
	\caption{Data and model predicted proportions of door choices in
	the IGT.}
\end{center}
\end{figure}

The initial preference for B and D choices confirms the theory
expressed in Huizenga et al. (2007) that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices reprensents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D reponses generate equal
profits in the long run.

@@

@@

There are different accounts of probabilistic category learning.
According to instance or exemplar learning theories, participants
learn by storing each encountered cue-outcome pairing.  When presented
with a cue pattern, these exemplars are retrieved from memory, and
weighted according to their similarity to the probe cue pattern, to
form a classification.  According to associative theories,
participants gradually learn by gradually associating the individual
cues (or cue patterns in configural learning) to the outcomes.  In
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  Gluck, Shohamy and Myers
(2002) proposed a number of such rules (or strategies).  A main
difference between these is whether responses are based on the
presence/absence of a single cue, or whether responses are based on
cue patterns.  Gluck et al.  formulated all strategies in a
deterministic and optimal manner (e.g., the multi-cue strategy
corresponded to giving the optimal response to each cue pattern).
Meeter et al.  allowed for probabilistic responding (a small
probability of giving the non-optimal response).

Alternative non-strategy based analyses of the WPT (Lagnado et al,
Speekenbrink et al) have estimated response strategies by logistic
regression, allowing the regression coefficients to change over time.

%associative, rule-based.
Here, we analyze the behavior of a single individual performing the
WPT for 200 trials.  We chose to analyse the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model).  We are particularly interested
in evidence for strategy switching and whether a DMM can recover a
strategy model in line with Gluck et al.  (2002).

@@

To reduce the degrees of freedom, and improve parameter estimates, we
implemented constraints to force state 1 into a single cue strategy
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue strategy (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue strategy was somewhat different than
described by Gluck et al.  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue strategy were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue strategy for the first 60 trials, and then
switched to the multi-cue strategy.

@@

References

Bechara, A., Damasio, A. R., Damasio, H., \& Anderson, S. W.(1994).
Insensitivity to future consequences following damage to human
prefrontal cortex.  Cognition, 50(1Ð3), 7Ð15.

Chambers, J. M. (1998). Programming with Data: A Guide to the S Lan-
guage. New York: Springer-Verlag.

Crone, E. A.,  \& van der Molen, M. W. (2004).  Developmental changes in
real life decision making: Performance on a gambling task previously
shown to depend on the ventromedial prefrontal cortex.  Developmental
Neuropsychology, 25(3), 251-279.

Dunn, B. D., Dalgleish, T.,  \& Lawrence, A. D. (2006).  The somatic
marker hypothesis: A critical evaluation.  Neuroscience and
Biobehavioral Reviews, 30(2), 239-271.

Gluck, M. A., Shohamy, D., \& Myers, C. (2002). How do people solve the
weather prediction task?: Individual variability in strategies for
probabilistic category learning. Learning \& Memory, 9, 408-418.

Huizenga, H. M., Crone, E. A., \& Jansen, B. R. J. (2007).
Decision-making in healthy children, adolescents and adults explained
by the use of increasingly complex proportional reasoning rules.
Developmental Science, 10(6), 814-825.

Knowlton, B. J., Squire, L. R., \& Gluck, M. A. (1994).
Probabilistic classification learning in amnesia.
Learning \& Memory, 1 , 106-120.

Siegler, R. S. (1981).  Developmental sequences within and between
concepts.  Monographs of the Society for Research in Child
Development, 46(2, Serial No.  189).

Visser, Raijmakers, \& Van der Maas (2008). Dynamics book chapter.

@@

\bibliography{all,ingmar}

>>>>

% phenomena of discrete change in development/learning
Piagetian developmental theory (REFERENCE??)  assumes step-wise
changes in the strategies that children apply in all kinds of tasks
such as the conservation learning and the balance scale task.
\citet{Maas1992} developed a catastrophe model to describe phase
transitions in learning and developmental processes.  They applied the
catastrophe model to learning in the conservation of liquid task
(REFERENCE??)  in which children have to judge relative volumes of
liquid in glasses of different heights and widths.  Young children
tend to ignore the width dimension and hence always choose the glass
with the highest level of liquid (REFERENCE??).  \citet{Maas1992}
showed that there is a sudden transition to a new strategy in which
the children also take the width of the glasses into account when
judging the volume of liquids.

% balance scale rules
\citet{Jansen2001} applied the catastrophe model to
development of strategies on the balance scale task (Siegler, 1981).
In the balance scale task participants have to judge which side of a
balance goes down when the number of weights and their distances to
the fulcrum are varied over trials.  Younger children tend to ignore
the distance dimension in this task, and instead focus solely on the
number of weights on each side of the fulcrum.  This strategy for
solving balance scale items is called Rule 1 \cite{Siegler1981}.  Older
children include the distance dimension in determining their response
to balance scale problems; however, they only do so when the weight
dimension does not differ between the sides of the balance, i.e., when
the number of weights is equal on both sides of the balance scale.
This strategy is called Rule 2 \cite{Siegler1981}.

% hysteresis on the balance scale
\citet{Jansen2001} found clear evidence for stage-wise
transitions between Rule 1 and Rule 2 by testing criteria that were
derived from the catastrophe model.  In particular, they found bimodal
test scores and inaccessibility.  The latter means that there are no
in-between strategies: children apply either Rule 1 or Rule 2 and
there is no in-between option.  \citet{Jansen2001} also found
evidence for hysteresis: the phenomenon that switching between
strategies is assymetric.  Children can switch from Rule 1 to Rule 2
and back, but this occurs at different trials.  In particular, if the
distance dimension in the balance scale problems is made more salient
by increasing the distance difference between weights on either side
of the balance scale, children may switch from Rule 1 to Rule 2.  If
subsequently the distance difference is decreased again, children may
switch back to using Rule 1.  Hysteresis is the phenomenon that this
switch back occurs at a different value of the control variable, in
this case the distance difference.

% conditioning and addiction research
Also in animal learning and conditioning, evidence is found for sudden
changes in response behavior \citet{Gallistel2004}.  In particular,
in their study, evidence was found for sudden onset of learning: at
the start of the learning experiment, the pigeons did not learn
anything and performance was stable; after a number of trials,
learning kicks in and there are large increases in performance.  The
interest here is in modeling the distribution of onset times: that is,
the trials at which learning suddenly takes off.  A similar interest
in process onset times is found in addiction research.  For example,
(REFERENCE) study the age at which children start using alcohol and
how this related to eventual outcomes in terms of addiction.

% discrimination and categorization learning
Sudden transitions in learning are also observed in simple
discrimination learning paradigms in which participants learn to
discriminate a number of stimuli based on a single dimension such as
form or color.  This kind of learning is referred to as all-or-none
learning or concept identification learning.  \citet{Raijmakers2001}
found evidence for different strategies applied by children when faced
with such a learning task.  \citet{Schmittmann2006} reanalyzed their
data using hidden Markov models to show that both strategies are
characterized by sudden transitions in the learning process.

% latent Markov versus time series analyses/outline
In above mentioned applications, the data consist mostly of a few
repeated measurements administered to large groups of participants.
The focus in the current chapter is rather on data that consist of
many repeated measurements, or time series, observed in only a few
participants or even just a single participant.  For example,
\citet{Visser2009} analyzed data from a single participant in an
experimental task that manipulates the trade-off between speed and
accuracy.  The data consisted of three time series with each around
150 repeated measurements of both reaction time and accuracy.  Below
we provide examples of analyzing time series from single participants
from two experiments; one from the Iowa Gambling Task and one from the
weather prediction task.  The interest in these tasks is to show that
participants develop different strategies over time in responding to
the stimuli and that the transition from one strategy to the next is a
discrete event.  Before providing these illustrations, below we give a
formalization of dependent mixture models and a brief overview of the
DepmixS4 package that was developed to specify and fit such models.

@@

% what is the IGT?
The Iowa gambling task (IGT) is an experimental paradigm designed to
mimic real-life decision-making situations \cite{Bechara1994}, in the
way that it factors uncertainty, reward and punishment
\cite{Dunn2006}.  The task requires the selection of cards from four
decks.  Each deck is characterized by a certain amount of gain
(delivered on each draw), frequency of loss, and amount of loss.  Two
decks (A and B) yield consistently high rewards, but also high,
probabilistic penalties and are both (equally) disadvantageous in the
long run.  The other two decks (C and D) yield consistently smaller
rewards, but also low, probabilistic penalties and are both (equally)
advantageous in the long run.  It is assumed that the ventromedial
prefrontal cortex (VMPFC) is active in the IGT as VMPC patients show
impaired task performance.  Their preference for the decks with
immediate high rewards indicates ``myopia for the future''.

% what is the HDT? developmental trends and relevance?
\cite{Crone2004} designed a developmentally appropriate analogue of
the IGT, the Hungry Donkey Task (HDT), with a similar win and loss
schedule although the abolute amounts were redcuced by a factor of 25.
The HDT is a pro-social game inviting the player to assist a hungry
donkey to collect as many apples as possible, by opening one of four
doors.  Again, doors A and B are characterized by a high constant gain
(10 apples), whereas doors C and D deliver a low constant gain (2
apples).  At doors A and C, a loss of 50 apples (A) or 10 apples (C)
is delivered in 50\% of the trials.  For doors B and D, frequency of
loss is only 10\%.  The median loss of doors B and D is 10 and 2,
respectively.  \cite{Crone2004} administered the HDT to children from
four age groups (6-9, 10-12, 13-15, and 18-25 year-olds) and concluded
that children also fail to consider future consequences.

% strategic reanalysis of the HDT: different strategies found by
% finite mixture analysis
A reanalysis of this dataset \cite{Huizenga2007} indicated that
participants might solve the task by sequentially considering the
three dimensions (constant gain, frequency of loss, and amount of
loss) in order to choose a door.  Most youngest children in the
dataset seem to focus on the dominant dimension in the task, frequency
of loss, resulting in equal preference for doors B and D. Older
participants seem to use a two-dimensional rule where participants
first focus on the frequency of loss and then consider amount of loss,
resulting in a preference for door D. A third very small subgroup
seems to use an integrative rule where participants combine all three
dimensions in the appropriate way.  Participants using the integrative
rule pick cards from doors C and D, which are advantageous in the long
run.

@@

We chose to analyze two participants that used different strategies at
the end of the task as analyzed in the manner proposed by
\citet{Huizenga2007}.  The first participants' data were best
described by a 4-state model.  This model's transition matrix is:
$$
\mat{A} = \begin{pmatrix} 
				0.64 & 0.33 & 0.00 & 0.03 \\
				0.10 & 0.89 & 0.01 & 0.00 \\
				0.05 & 0.00 & 0.95 & 0.00 \\
				0.00 & 0.00 & 0.00 & 1.00
		  \end{pmatrix}
$$
As can be seen from the transition matrix the final state is
absorbind.  States 2 and 3 are also fairly stable with high diagonal
probabilities of 0.89 and 0.95 respectively.  The response parameters
are presented in Table~ref{tab:igt4} below.

@@

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/post4.pdf}
	\label{fig:post4}
	\caption{Posterior state sequence for the 4 state model.}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
	\includegraphics[width=\textwidth]{graphs/igt4.pdf}
	\label{fig:igt4}
	\caption{Data and model predicted proportions of door choices in
	the IGT.}
\end{center}
\end{figure}

The initial preference for B and D choices confirms the theory
expressed in \citet{Huizenga2007} that frequency of loss is the
dominant dimension in the IGT. The final states with mostly C 
choices reprensents one of the optimal strategies, which consists of
both C and D responses. Note that both C and D reponses generate equal
profits in the long run.

@@

\nocite{Knowlton1994}

@@

There are different accounts of probabilistic category learning.
According to instance or exemplar learning theories, participants
learn by storing each encountered cue-outcome pairing.  When presented
with a cue pattern, these exemplars are retrieved from memory, and
weighted according to their similarity to the probe cue pattern, to
form a classification.  According to associative theories,
participants gradually learn by gradually associating the individual
cues (or cue patterns in configural learning) to the outcomes.  In
rule-learning, participants are taken to extract rules by which to
categorize the different cue patterns.  \cite{Gluck2002} proposed a
number of such rules (or strategies).  A main difference between these
is whether responses are based on the presence/absence of a single
cue, or whether responses are based on cue patterns.  Gluck et al.
formulated all strategies in a deterministic and optimal manner (e.g.,
the multi-cue strategy corresponded to giving the optimal response to
each cue pattern).  Meeter et al.  allowed for probabilistic
responding (a small probability of giving the non-optimal response).

Alternative non-strategy based analyses of the WPT
\cite{Lagnado??,Speekenbrink??} have estimated response strategies by
logistic regression, allowing the regression coefficients to change
over time.

%associative, rule-based.
Here, we analyze the behavior of a single individual performing the
WPT for 200 trials.  We chose to analyse the ``average'' participant
(the participant with performance closest to the group average) in a
large unpublished dataset.  We let each state be characterized by a
GLM with a Binomial distributed response and logistic link function
(i.e., a logistic regression model).  We are particularly interested
in evidence for strategy switching and whether a DMM can recover a
strategy model in line with \cite{Gluck2002}.

@@

To reduce the degrees of freedom, and improve parameter estimates, we
implemented constraints to force state 1 into a single cue strategy
(fixing the coefficients of the remaining three cues to 0) and state 2
in a multi-cue strategy (forcing the intercept to 0).  These
restrictions resulted in a better AIC value of AIC=185.24 (df=7).
Interestingly, the single cue strategy was somewhat different than
described by \citet{Gluck2002}  Parameter estimates indicated relatively
more consistent predictions of ``rain'' in the absence of cue 1
($Pr(\text{sun}) = 0.22$) and more inconsistent predictions of ``sun''
in the presence of cue 1 ($Pr(\text{sun}) = 0.60$).  The cue weights
of the multi-cue strategy were in the direction of the optimal
weights.  The Viterbi state sequence indicated that the participant
used the single cue strategy for the first 60 trials, and then
switched to the multi-cue strategy.

@@

@@

\bibliography{all,ingmar,individual}
